---
title: "Game, dataSet, Match!"
author: "Evan O'Neil"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(ggplot2)
library(knitr)
library(DT)
library(kableExtra)
library(scales)
library(lubridate)
library(viridis)
library(leaflet)  # For interactive maps
library(countrycode)  # For country code conversion
library(htmltools)  # For HTML widgets
library(plotly)  # For interactive plots
library(broom)  # For cleaning up statistical output

# Set theme for consistent visualization
theme_set(theme_minimal())
```

# 1. Introduction {.tabset}

## Overview
This document is a draft of an Exploratory Data Analysis of all ATP Tennis matches from the 2024 season. The dataset contains information about professional tennis matches played during the 2024 ATP Tour, including match outcomes, player statistics, tournament details, and more.

## Data Source
A note, I changed my data source, though the spirit of the data remains the same from my proposal. In my exploratory work I noticed inconsistencies in the statistical output adn dug deeper to find the dataset was incomplete, it stopped before the U.S. Open. While a setback, this represents one of the important outcomes of exploratory work when combined with subject matter expertise.

I searched for the sources of the data and realized it was all coming from Tennis Abstract, the leading name for tennis data. With a littile more digging I saw that all this data and more was available on Jeff Sackmann's (the lead organizer of Tennis Abstract) GitHub page. I saw that not only was the complete data for the 2023 season available, but also the complete data for the 2024 season is available too. I chose to switch to the more recent data. The structure is the same, but with the most recent data I thought I could ultimately transform this into content for my tennis analysis newsletter and 2024 data would be of more interest to my readers than 2023. 

# 2. Data Loading and Initial Exploration {.tabset}

## Data Loading
```{r load-data}
# Function to download and load data directly from a GitHub URL
get_github_data <- function(github_url) {
  # Convert GitHub URL to raw URL
  raw_url <- github_url
  raw_url <- gsub("github\\.com", "raw.githubusercontent.com", raw_url)
  raw_url <- gsub("/blob/", "/", raw_url)
  
  # Create a temporary file
  temp_file <- tempfile(fileext = ".csv")
  
  # Download the file
  message("Downloading data from GitHub...")
  download_success <- tryCatch({
    download.file(raw_url, temp_file, mode = "wb", quiet = TRUE)
    TRUE
  }, error = function(e) {
    message("Error downloading the file: ", e$message)
    FALSE
  })
  
  # Load the data if download was successful
  if (download_success && file.exists(temp_file)) {
    message("Download successful, loading data...")
    data <- read.csv(temp_file)
    return(data)
  } else {
    message("Failed to download or load the data.")
    return(NULL)
  }
}

tennis_data <- get_github_data("https://github.com/JeffSackmann/tennis_atp/blob/master/atp_matches_2024.csv")

# Check if data loaded successfully
if (!is.null(tennis_data)) {
  # Show first few rows
  head(tennis_data)
  
  # Get dimensions
  message("Dataset dimensions: ", nrow(tennis_data), " rows × ", ncol(tennis_data), " columns")
}
```

## 2.1 Dataset Overview

```{r dataset-overview}
# Display the first few rows as a DT table
datatable(head(tennis_data, 10),
          options = list(scrollX = TRUE, 
                         autoWidth = TRUE,
                         pageLength = 5),
          caption = "First 10 Rows of ATP Tennis 2023 Dataset") %>%
  formatStyle(columns = colnames(tennis_data), fontSize = '90%')
```

## 2.2 Column Names and Data Types

```{r column-info}
# List all column names
col_names <- colnames(tennis_data)

# Create a data frame to explain column names in direct terms using my vast tennis knowledge
column_explanations <- data.frame(
  Column = c(
    "tourney_id", "tourney_name", "surface", "draw_size", "tourney_level",
    "winner_id", "winner_seed", "winner_name", "winner_hand", "winner_ht", "winner_ioc", "winner_age",
    "loser_id", "loser_seed", "loser_name", "loser_hand", "loser_ht", "loser_ioc", "loser_age",
    "score", "best_of", "round", "minutes",
    "w_ace", "w_df", "w_svpt", "w_1stIn", "w_1stWon", "w_2ndWon", "w_SvGms", "w_bpSaved", "w_bpFaced",
    "l_ace", "l_df", "l_svpt", "l_1stIn", "l_1stWon", "l_2ndWon", "l_SvGms", "l_bpSaved", "l_bpFaced",
    "winner_rank", "winner_rank_points", "loser_rank", "loser_rank_points"
  ),
  Description = c(
    "Unique tournament identifier", "Tournament name", "Playing surface type (Hard, Clay, Grass, Carpet)", 
    "Number of players in the tournament draw", "Tournament tier (G=Grand Slam, M=Masters 1000, A=ATP 500/250, D=Davis Cup, F=Tour Finals)",
    "Winner's player ID", "Winner's seeding in the tournament", "Winner's full name", "Winner's playing hand (R=Right, L=Left)", 
    "Winner's height in cm", "Winner's country code", "Winner's age in years",
    "Loser's player ID", "Loser's seeding in the tournament", "Loser's full name", "Loser's playing hand (R=Right, L=Left)", 
    "Loser's height in cm", "Loser's country code", "Loser's age in years",
    "Match score (sets)", "Maximum number of sets (3 or 5)", "Tournament round", "Match duration in minutes",
    "Winner's ace count", "Winner's double fault count", "Winner's service points played", 
    "Winner's first serves in", "Winner's first serve points won", "Winner's second serve points won",
    "Winner's service games played", "Winner's break points saved", "Winner's break points faced",
    "Loser's ace count", "Loser's double fault count", "Loser's service points played", 
    "Loser's first serves in", "Loser's first serve points won", "Loser's second serve points won",
    "Loser's service games played", "Loser's break points saved", "Loser's break points faced",
    "Winner's ATP ranking", "Winner's ATP ranking points", "Loser's ATP ranking", "Loser's ATP ranking points"
  )
)

# Display column explanations in a wonderful table
library(DT)
datatable(column_explanations, 
          options = list(scrollX = TRUE, 
                         autoWidth = TRUE,
                         pageLength = 15),
          caption = "Summary Statistics for Key Numeric Variables",
          rownames = FALSE) 
```

## 2.3 Summary Statistics 

```{r summary-stats}
# Get all numeric columns for comprehensive summary
numeric_cols <- sapply(tennis_data, is.numeric)
numeric_col_names <- names(tennis_data)[numeric_cols]

# Function to calculate summary statistics for a numeric column
get_column_summary <- function(data, column) {
  if(column %in% colnames(data)) {
    values <- data[[column]]
    values <- values[!is.na(values)]
    
    if(length(values) > 0) {
      return(data.frame(
        Column = column,
        Min = min(values, na.rm = TRUE),
        Q1 = quantile(values, 0.25, na.rm = TRUE),
        Median = median(values, na.rm = TRUE),
        Mean = mean(values, na.rm = TRUE),
        Q3 = quantile(values, 0.75, na.rm = TRUE),
        Max = max(values, na.rm = TRUE),
        Missing = sum(is.na(data[[column]])),
        Missing_Pct = round(sum(is.na(data[[column]])) / nrow(data) * 100, 1)
      ))
    }
  }
  return(NULL)
}

# Collect summary statistics for each numeric column
summary_list <- lapply(numeric_col_names, function(col) get_column_summary(tennis_data, col))
summary_df <- do.call(rbind, summary_list)

# Format dates using base R
if("tourney_date" %in% colnames(summary_df)) {
  if(class(tennis_data$tourney_date)[1] == "Date") {
    summary_df$Min[summary_df$Column == "tourney_date"] <- as.character(as.Date(summary_df$Min[summary_df$Column == "tourney_date"], origin = "1970-01-01"))
    summary_df$Median[summary_df$Column == "tourney_date"] <- as.character(as.Date(summary_df$Median[summary_df$Column == "tourney_date"], origin = "1970-01-01"))
    summary_df$Mean[summary_df$Column == "tourney_date"] <- as.character(as.Date(summary_df$Mean[summary_df$Column == "tourney_date"], origin = "1970-01-01"))
    summary_df$Q1[summary_df$Column == "tourney_date"] <- as.character(as.Date(summary_df$Q1[summary_df$Column == "tourney_date"], origin = "1970-01-01"))
    summary_df$Q3[summary_df$Column == "tourney_date"] <- as.character(as.Date(summary_df$Q3[summary_df$Column == "tourney_date"], origin = "1970-01-01"))
    summary_df$Max[summary_df$Column == "tourney_date"] <- as.character(as.Date(summary_df$Max[summary_df$Column == "tourney_date"], origin = "1970-01-01"))
  }
}

# Display using DT
datatable(summary_df, 
          options = list(scrollX = TRUE, 
                         pageLength = 15,
                         dom = 'ftip'),
          caption = "Summary Statistics for All Numeric Variables",
          rownames = FALSE)
```

```{r numeric-small-multiples-ggplot, fig.width=12, fig.height=15, warning=FALSE, message=FALSE}
# Get all numeric columns
numeric_cols <- names(tennis_data)[sapply(tennis_data, is.numeric)]

# Select a subset of numeric columns (exclude IDs and dates)
important_numeric_cols <- numeric_cols[!(numeric_cols %in% c("tourney_id", "match_num", "tourney_date", "winner_id", "loser_id"))]

# Create a long-format data frame for plotting
numeric_data_long <- tennis_data %>%
  select(all_of(important_numeric_cols[1:min(16, length(important_numeric_cols))])) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Calculate summary statistics for each variable
stats_summary <- numeric_data_long %>%
  group_by(variable) %>%
  summarise(
    mean_val = mean(value, na.rm = TRUE),
    median_val = median(value, na.rm = TRUE),
    min_val = min(value, na.rm = TRUE),
    max_val = max(value, na.rm = TRUE)
  )

# Create a function to add labels to each facet
label_facets <- function(orig_var, stats_df) {
  # Get stats for this variable
  var_stats <- stats_df %>% filter(variable == orig_var)
  
  # Make a nice variable name
  nice_name <- gsub("_", " ", orig_var)
  nice_name <- gsub("([[:lower:]])([[:upper:]])", "\\1 \\2", nice_name)
  nice_name <- tools::toTitleCase(nice_name)
  
  # Return the formatted label
  return(paste0(
    nice_name, "\n",
    "Mean: ", round(var_stats$mean_val, 1), "\n",
    "Median: ", round(var_stats$median_val, 1), "\n",
    "Min: ", round(var_stats$min_val, 1), "\n",
    "Max: ", round(var_stats$max_val, 1)
  ))
}

# Create the plot
ggplot(numeric_data_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "darkblue", alpha = 0.7) +
  geom_vline(data = stats_summary, aes(xintercept = mean_val), 
             color = "red", linetype = "dashed", size = 0.8) +
  geom_vline(data = stats_summary, aes(xintercept = median_val), 
             color = "blue", linetype = "solid", size = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 4,
             labeller = labeller(variable = function(x) sapply(x, label_facets, stats_summary))) +
  labs(
    title = "Distribution of Numeric Variables",
    subtitle = "With mean (red dashed) and median (blue solid) lines",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 7),
    panel.spacing = unit(1, "lines")
  )
```

## 2.4 Missing Values Analysis

```{r missing-values}
# Calculate missing values in each column
missing_values <- colSums(is.na(tennis_data))

# Show columns with missing values
missing_df <- data.frame(
  Column = names(missing_values),
  Missing_Count = missing_values,
  Missing_Percent = round(missing_values / nrow(tennis_data) * 100, 2)
) %>%
  arrange(desc(Missing_Count))

# Display columns with any missing values using DT
datatable(missing_df %>% filter(Missing_Count > 0), 
          options = list(pageLength = 15,
                         dom = 'ltip',
                         order = list(list(2, 'desc'))),
          caption = "Columns with Missing Values") %>%
  formatStyle('Missing_Percent',
              background = styleColorBar(c(0, 100), 'lightblue'),
              backgroundSize = '100% 90%',
              backgroundRepeat = 'no-repeat',
              backgroundPosition = 'center')
```

# 3. Data Cleaning and Preparation {.tabset}

## Cleaning Steps
```{r data-cleaning}
# Create a clean working copy of the dataset
atp_clean <- tennis_data %>%
  # Convert tournament date to actual date format if it's not already
  mutate(
    tourney_date_str = as.character(tourney_date),
    # Format assuming YYYYMMDD format
    tourney_date = if(nchar(tourney_date_str[1]) == 8) {
      ymd(tourney_date_str)
    } else {
      tourney_date  
    }
  ) %>%
  select(-tourney_date_str)  # Remove temporary column

# Extract year, month from tournament date if it was converted successfully
if(is.Date(atp_clean$tourney_date[1])) {
  atp_clean <- atp_clean %>%
    mutate(
      year = year(tourney_date),
      month = month(tourney_date)
    )
}

# Add calculated metrics where possible
atp_clean <- atp_clean %>%
  mutate(
    # Only calculate percentages when denominators exist and are not zero
    winner_1st_serve_pct = ifelse(!is.na(w_svpt) & !is.na(w_1stIn) & w_svpt > 0, 
                                 w_1stIn / w_svpt * 100, NA),
    winner_1st_serve_won_pct = ifelse(!is.na(w_1stIn) & !is.na(w_1stWon) & w_1stIn > 0,
                                     w_1stWon / w_1stIn * 100, NA),
    loser_1st_serve_pct = ifelse(!is.na(l_svpt) & !is.na(l_1stIn) & l_svpt > 0,
                                l_1stIn / l_svpt * 100, NA),
    # Create categorical Round variable
    round_category = case_when(
      round == "F" ~ "Final",
      round == "SF" ~ "Semi-Final",
      round == "QF" ~ "Quarter-Final",
      round == "R16" ~ "Round of 16",
      round == "R32" ~ "Round of 32",
      round == "R64" ~ "Round of 64",
      round == "R128" ~ "Round of 128",
      TRUE ~ round
    )
  )

# List of ATP 500 tournaments for proper classification
atp_500_tournaments <- c("Acapulco", "Barcelona", "Dubai", "Halle", "Hamburg", 
                         "Queen's Club", "Rotterdam", "Washington", "Rio De Janeiro")

# Add tournament level classification
atp_clean <- atp_clean %>%
  mutate(
    # Add 500/250 tournament level distinction as "A" classification is insufficient
    tourney_level_detail = case_when(
      tourney_level == "G" ~ "Grand Slam",
      tourney_level == "M" ~ "Masters 1000",
      tourney_level == "F" ~ "Tour Finals",
      tourney_level == "D" ~ "Davis Cup",
      tourney_level == "O" ~ "Olympics", 
      tourney_level == "A" & str_detect(tourney_name, paste(atp_500_tournaments, collapse = "|")) ~ "ATP Tour 500",
      tourney_level == "A" ~ "ATP Tour 250",
      TRUE ~ "Other"
    )
  )

# Create a table showing tournament level classification results
tourney_level_counts <- atp_clean %>%
  group_by(tourney_level_detail) %>%
  summarise(
    Tournaments = n_distinct(tourney_id),
    Matches = n()
  ) %>%
  arrange(desc(Matches))

# Display table with tournament level counts
datatable(tourney_level_counts,
          options = list(dom = 't', 
                         ordering = FALSE),
          caption = "Tournament Levels in the Dataset")

# Show a quick summary of the derived variables
derived_vars <- atp_clean %>%
  summarise(
    `Avg Winner 1st Serve %` = round(mean(winner_1st_serve_pct, na.rm = TRUE), 1),
    `Avg Winner 1st Serve Won %` = round(mean(winner_1st_serve_won_pct, na.rm = TRUE), 1),
    `Avg Loser 1st Serve %` = round(mean(loser_1st_serve_pct, na.rm = TRUE), 1),
    `Total Tournaments` = n_distinct(tourney_id),
    `Total Players` = n_distinct(c(winner_name, loser_name)))

# Display derived variables summary
datatable(derived_vars,
          options = list(dom = 't',
                         ordering = FALSE,
                         paging = FALSE),
          caption = "Summary of Derived Variables")
```

# 4. Data Analysis {.tabset}

## 4.1 Player Analysis {.tabset .tabset-pills}

### 4.1.1 Geographic Analysis {.tabset}

#### Data Table
```{r player-countries-map-fixed, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
# Install and load required packages
if (!require("countrycode")) {
  install.packages("countrycode")
  library(countrycode)
} else {
  library(countrycode)
}

# Extract unique players with their countries
unique_players <- rbind(
  atp_clean %>% 
    select(player_id = winner_id, player_name = winner_name, ioc = winner_ioc) %>%
    distinct(),
  atp_clean %>% 
    select(player_id = loser_id, player_name = loser_name, ioc = loser_ioc) %>%
    distinct()
) %>%
  distinct(player_id, .keep_all = TRUE) %>%
  filter(!is.na(player_id) & !is.na(ioc))

# Count players per country
country_counts <- table(unique_players$ioc)
country_df <- data.frame(
  ioc = names(country_counts),
  count = as.numeric(country_counts)
)

# Convert IOC codes to ISO codes for mapping
country_df$iso3c <- countrycode(country_df$ioc, "ioc", "iso3c")
country_df$country_name <- countrycode(country_df$ioc, "ioc", "country.name")

# Create world map data
world_map <- map_data("world")

# Fix specific countries with known mapping issues
# Fix for the smallest of the continued issues that Russia presents
if ("RUS" %in% country_df$ioc) {
  # Get Russia's player count
  russia_count <- country_df$count[country_df$ioc == "RUS"]
}

# Fix for USA
if ("USA" %in% country_df$ioc) {
  country_df$country_name[country_df$ioc == "USA"] <- "USA"
}

# Fix for UK
if ("GBR" %in% country_df$ioc) {
  country_df$country_name[country_df$ioc == "GBR"] <- "UK"
}

# Join the country counts with map data
map_data <- left_join(
  world_map,
  country_df,
  by = c("region" = "country_name")
)

# Explicitly set Russia's count in the map data
map_data$count[map_data$region == "Russia"] <- russia_count

# Fill NA values with 0
map_data$count[is.na(map_data$count)] <- 0

# Create the chloropleth map
ggplot(map_data, aes(x = long, y = lat, group = group, fill = count)) +
  geom_polygon(color = "white", size = 0.1) +
  scale_fill_viridis_c(
    name = "Player Count",
    option = "plasma",
    direction = -1,
    limits = c(0, max(country_df$count)),
    breaks = c(0, 5, 10, 15, 20, 30, 35)
  ) +
  labs(
    title = "Geographic Distribution of ATP Tennis Players (2024)",
    subtitle = paste0("Based on unique players per country, total of ", nrow(unique_players), " players"),
    caption = "Note: Countries with no players shown in yellow"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

#### Interactive Map
```{r player-map-leaflet, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
# Extract unique players with their countries
unique_players <- rbind(
  atp_clean %>% 
    select(player_id = winner_id, player_name = winner_name, ioc = winner_ioc) %>%
    distinct(),
  atp_clean %>% 
    select(player_id = loser_id, player_name = loser_name, ioc = loser_ioc) %>%
    distinct()
) %>%
  distinct(player_id, .keep_all = TRUE) %>%
  filter(!is.na(player_id) & !is.na(ioc))

# Count players per country
country_counts <- table(unique_players$ioc)
country_df <- data.frame(
  ioc = names(country_counts),
  count = as.numeric(country_counts)
)

# Convert IOC codes to ISO3 codes for mapping
country_df$iso3c <- countrycode(country_df$ioc, "ioc", "iso3c")
country_df$country_name <- countrycode(country_df$ioc, "ioc", "country.name")

# Handle missing conversions or special cases
country_df$country_name[country_df$ioc == "USA"] <- "United States"
country_df$country_name[country_df$ioc == "GBR"] <- "United Kingdom"
country_df$iso3c[country_df$ioc == "RUS"] <- "RUS" # Ensure Russia has a code

# Create a world map with Leaflet
# Get world country polygons from the rworldmap package
if (!require("rworldmap")) {
  install.packages("rworldmap")
  library(rworldmap)
}

# Get world map data
world_map_data <- rworldmap::getMap(resolution = "low")

# Create a color palette
pal <- colorNumeric(
  palette = "plasma",
  domain = country_df$count,
  reverse = TRUE
)

# Initialize the leaflet map
leaflet(world_map_data) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  setView(lng = 0, lat = 30, zoom = 2) %>%  # Set initial view
  # Add countries colored by player count
  addPolygons(
    fillColor = ~pal(country_df$count[match(NAME, country_df$country_name)]),
    weight = 1,
    opacity = 1,
    color = "white",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0(
      NAME, ": ", 
      ifelse(is.na(country_df$count[match(NAME, country_df$country_name)]), 
             "0", 
             country_df$count[match(NAME, country_df$country_name)]),
      " players"
    ),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  # Add a legend
  addLegend(
    position = "bottomright",
    pal = pal,
    values = country_df$count,
    title = "Player Count",
    opacity = 0.7
  )
```

#### Player Distribution Bar Chart
```{r player-nationality-bar, fig.width=10, fig.height=8}
# Get the top 25 countries by player count
top_countries <- country_df %>%
  arrange(desc(count)) %>%
  head(25)

# Create a bar chart
ggplot(top_countries, aes(x = reorder(ioc, count), y = count, fill = count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = count), hjust = -0.2) +
  scale_fill_viridis_c(direction = -1) +
  labs(
    title = "Top 25 Countries by Number of ATP Tennis Players (2023)",
    subtitle = "Based on unique player IDs per country",
    x = "Country (IOC Code)",
    y = "Number of Players"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  coord_flip()
```

The colors should match up between the chart and the map. I was having difficulties matching them, this will be fixed in the final version.

### 4.1.2 Box plot of player age
```{r player-age-boxplot, fig.width=8, fig.height=6}
# Extract player ages
player_ages <- c(atp_clean$winner_age, atp_clean$loser_age)
player_ages <- player_ages[!is.na(player_ages)]

# Create data frame for plotting
age_df <- data.frame(
  age = player_ages,
  category = "Player Age"
)

# Generate summary statistics for reference
age_summary <- data.frame(
  Min = min(player_ages, na.rm = TRUE),
  Q1 = quantile(player_ages, 0.25, na.rm = TRUE),
  Median = median(player_ages, na.rm = TRUE),
  Mean = mean(player_ages, na.rm = TRUE),
  Q3 = quantile(player_ages, 0.75, na.rm = TRUE),
  Max = max(player_ages, na.rm = TRUE),
  SD = sd(player_ages, na.rm = TRUE)
)

# Create the box plot
ggplot(age_df, aes(x = category, y = age)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.1, color = "darkblue") +
  labs(
    title = "Distribution of ATP Tennis Player Ages (2024)",
    subtitle = paste0("Mean age: ", round(mean(player_ages, na.rm = TRUE), 1), 
                     " years | Median age: ", round(median(player_ages, na.rm = TRUE), 1), " years"),
    y = "Age (years)",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )

# Display summary statistics using DT
datatable(age_summary, 
          caption = "Summary Statistics for Player Age",
          options = list(dom = 't', 
                         pageLength = 1,
                         scrollX = TRUE,
                         columnDefs = list(list(
                           className = 'dt-center', 
                           targets = "_all"))),
          rownames = FALSE) %>%
  formatRound(columns = names(age_summary), digits = 1)
```

### 4.1.3 Box plot of player height
#### Or a tale of Viacheslav Bielinskyi, the 2 foot, 330 pound player from Ukraine.
```{r player-height-boxplot, fig.width=8, fig.height=6}
# Extract player heights
player_heights <- c(atp_clean$winner_ht, atp_clean$loser_ht)
player_heights <- player_heights[!is.na(player_heights)]

# Create data frame for plotting
height_df <- data.frame(
  height = player_heights,
  category = "Player Height"
)

# Generate summary statistics for reference
height_summary <- data.frame(
  Min = min(player_heights, na.rm = TRUE),
  Q1 = quantile(player_heights, 0.25, na.rm = TRUE),
  Median = median(player_heights, na.rm = TRUE),
  Mean = mean(player_heights, na.rm = TRUE),
  Q3 = quantile(player_heights, 0.75, na.rm = TRUE),
  Max = max(player_heights, na.rm = TRUE),
  SD = sd(player_heights, na.rm = TRUE)
)

# Create the box plot
ggplot(height_df, aes(x = category, y = height)) +
  geom_boxplot(fill = "seagreen", alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.1, color = "darkgreen") +
  labs(
    title = "Distribution of ATP Tennis Player Heights (2024)",
    subtitle = paste0("Mean height: ", round(mean(player_heights, na.rm = TRUE), 1), 
                     " cm | Median height: ", round(median(player_heights, na.rm = TRUE), 1), " cm"),
    y = "Height (cm)",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )

# Display summary statistics using DT
datatable(height_summary, 
          caption = "Summary Statistics for Player Height (cm)",
          options = list(dom = 't', 
                         pageLength = 1,
                         scrollX = TRUE),
          rownames = FALSE) %>%
  formatRound(columns = names(height_summary), digits = 1)
```
This represents an interesting error. I thought the 71cm height of a player was an error, especially as I experienced the difficult competitive prospects of being 5'10", I was very impressed that a player who was under 3 feet could compete at a major event. Even more impressive was the fact that upon looking him up of the official ATP website, I saw that he weighs in at 330 pounds! To compete at this level, with these physical dimensions, all while your home country is being subjected to a brutal invasion is quite impressive. This is all from official documentation so I left it in, surely an organization as well funded as the ATP would have corrected any errors on this scale if they were true errors.

### 4.1.4 Bar chart of right handed vs left handed players
```{r player-handedness-barchart, fig.width=8, fig.height=6}
# Extract player handedness
player_hands <- c(atp_clean$winner_hand, atp_clean$loser_hand)
hand_counts <- table(player_hands)

# Create data frame for plotting
hand_df <- data.frame(
  hand = names(hand_counts),
  count = as.numeric(hand_counts)
)

# Recode for better labels
hand_df$hand_label <- factor(hand_df$hand, 
                             levels = c("R", "L", "U"), 
                             labels = c("Right-handed", "Left-handed", "Unknown"))

# Calculate percentages
hand_df$percentage <- hand_df$count / sum(hand_df$count) * 100

# Bar chart of handedness
ggplot(hand_df, aes(x = hand_label, y = count, fill = hand_label)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = paste0(count, " (", round(percentage, 1), "%)")),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Right-handed" = "#3366CC", 
                               "Left-handed" = "#FF9933", 
                               "Unknown" = "#CCCCCC")) +
  labs(
    title = "Distribution of ATP Tennis Player Handedness (2024)",
    subtitle = "Count and percentage of right-handed vs. left-handed players",
    y = "Number of Player Appearances",
    x = NULL,
    fill = "Handedness"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 12),
    legend.position = "bottom"
  )

# Display table using DT
datatable(hand_df[, c("hand_label", "count", "percentage")], 
          caption = "Player Handedness Distribution",
          options = list(dom = 't', 
                         pageLength = 3,
                         scrollX = TRUE),
          colnames = c("Handedness", "Count", "Percentage (%)"),
          rownames = FALSE) %>%
  formatRound(columns = "percentage", digits = 1) %>%
  formatStyle('percentage',
              background = styleColorBar(c(0, max(hand_df$percentage)), 'lightblue'),
              backgroundSize = '95% 80%',
              backgroundRepeat = 'no-repeat',
              backgroundPosition = 'center')
```



## 4.2 Tournament Analysis

```{r tournament-analysis}
# Analyze tournament levels with the detailed classification
if("tourney_level_detail" %in% colnames(atp_clean)) {
  # Count matches by detailed tournament level
  tourney_counts_detail <- atp_clean %>%
    group_by(tourney_level_detail) %>%
    summarise(
      tournaments = n_distinct(tourney_id),
      matches = n()
    ) %>%
    arrange(desc(matches))
  
  # Display results
  tourney_counts_detail
  
  # Create visualization with separate 500 and 250 levels
  ggplot(tourney_counts_detail, aes(x = reorder(tourney_level_detail, matches), 
                                   y = matches, fill = tourney_level_detail)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = matches), vjust = -0.5, color = "black", size = 3.5) +
    labs(title = "Number of Matches by Tournament Level",
         subtitle = "With ATP 500 and ATP 250 properly distinguished",
         x = "Tournament Level",
         y = "Number of Matches") +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  # You can also add this analysis of tournament levels by surface
  tourney_surface <- atp_clean %>%
    group_by(tourney_level_detail, surface) %>%
    summarise(matches = n()) %>%
    arrange(tourney_level_detail, desc(matches))
  
  # Create stacked bar chart of surfaces by tournament level
  ggplot(tourney_surface, aes(x = tourney_level_detail, y = matches, fill = surface)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Tournament Levels by Surface",
         x = "Tournament Level",
         y = "Number of Matches",
         fill = "Surface") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

# 5. Player Performance Analysis

```{r player-performance}
# Top players by wins
top_winners <- atp_clean %>%
  group_by(winner_name) %>%
  summarise(
    wins = n(),
    win_pct = n() / length(unique(c(winner_name, loser_name))) * 100,
    avg_rank = mean(winner_rank, na.rm = TRUE)
  ) %>%
  arrange(desc(wins)) %>%
  head(10)

# Display results with DT
datatable(top_winners,
          options = list(dom = 't',
                         ordering = TRUE),
          caption = "Top 10 Players by Number of Wins") %>%
  formatRound(columns = c("win_pct", "avg_rank"), digits = 1)

# Create visualization
ggplot(top_winners, aes(x = reorder(winner_name, wins), y = wins)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = wins), hjust = -0.2) +
  labs(title = "Top 10 Players by Number of Wins",
       x = "Player",
       y = "Number of Wins") +
  coord_flip() +
  theme_minimal()

# Calculate win/loss records for all players
all_players <- data.frame(
  player = c(atp_clean$winner_name, atp_clean$loser_name),
  result = c(rep("win", nrow(atp_clean)), rep("loss", nrow(atp_clean)))
)

player_records <- all_players %>%
  group_by(player) %>%
  summarise(
    matches = n(),
    wins = sum(result == "win"),
    losses = sum(result == "loss"),
    win_pct = round(wins / matches * 100, 1)
  ) %>%
  filter(matches >= 10) %>%  # Only include players with at least 10 matches
  arrange(desc(win_pct)) %>%
  head(10)

# Display top players by win percentage
datatable(player_records,
          options = list(dom = 't',
                         pageLength = 10),
          caption = "Top 10 Players by Win Percentage (min. 10 matches)")
```
Including only players with 15 matches was an entirely arbitrary number selection, I'll search for more substantive ways to make that decision.

# 6. Surface Analysis 

```{r surface-analysis}
# Match duration analysis by surface
if("minutes" %in% colnames(atp_clean) && "surface" %in% colnames(atp_clean)) {
  # Calculate average duration by surface
  duration_by_surface <- atp_clean %>%
    filter(!is.na(minutes)) %>%
    group_by(surface) %>%
    summarise(
      avg_duration = mean(minutes, na.rm = TRUE),
      median_duration = median(minutes, na.rm = TRUE),
      min_duration = min(minutes, na.rm = TRUE),
      max_duration = max(minutes, na.rm = TRUE),
      matches = n()
    ) %>%
    arrange(desc(avg_duration))
  
  # Display results with DT
  datatable(duration_by_surface,
            options = list(dom = 't',
                           ordering = TRUE),
            caption = "Match Duration by Surface") %>%
    formatRound(columns = c("avg_duration", "median_duration", "min_duration", "max_duration"), digits = 1)
  
  # Create visualization
  ggplot(duration_by_surface, aes(x = reorder(surface, avg_duration), y = avg_duration, fill = surface)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = round(avg_duration, 1)), vjust = -0.5, color = "black", size = 3.5) +
    labs(title = "Average Match Duration by Surface (minutes)",
         x = "Surface",
         y = "Average Duration (minutes)") +
    theme_minimal() +
    theme(legend.position = "none")
}

# Players with most matches on each surface
players_by_surface <- atp_clean %>%
  filter(!is.na(surface)) %>%
  group_by(player_name = winner_name, surface) %>%
  summarise(wins = n()) %>%
  bind_rows(
    atp_clean %>%
      filter(!is.na(surface)) %>%
      group_by(player_name = loser_name, surface) %>%
      summarise(losses = n())
  ) %>%
  group_by(player_name, surface) %>%
  summarise(
    matches = sum(wins, na.rm = TRUE) + sum(losses, na.rm = TRUE),
    wins = sum(wins, na.rm = TRUE),
    win_pct = round(wins / matches * 100, 1)
  ) %>%
  arrange(surface, desc(matches)) %>%
  group_by(surface) %>%
  slice_head(n = 5)

# Display players by surface
datatable(players_by_surface,
          options = list(pageLength = 15,
                         order = list(list(0, 'asc'), list(2, 'desc'))),
          caption = "Top 5 Players by Number of Matches on Each Surface")

# Analyze aces by surface
if(all(c("w_ace", "l_ace", "surface") %in% colnames(atp_clean))) {
  # Calculate average aces by surface
  aces_by_surface <- atp_clean %>%
    filter(!is.na(w_ace) & !is.na(l_ace)) %>%
    group_by(surface) %>%
    summarise(
      avg_total_aces = mean(w_ace + l_ace, na.rm = TRUE),
      matches = n()
    ) %>%
    arrange(desc(avg_total_aces))
  
  # Display results
  aces_by_surface
  
  # Create visualization
  ggplot(aces_by_surface, aes(x = reorder(surface, avg_total_aces), y = avg_total_aces, fill = surface)) +
    geom_bar(stat = "identity") +
    labs(title = "Average Aces per Match by Surface",
         x = "Surface",
         y = "Average Aces") +
    theme(legend.position = "none")
}

# Analyze first serve percentage by surface
if("winner_1st_serve_pct" %in% colnames(atp_clean) && "surface" %in% colnames(atp_clean)) {
  # Calculate average first serve percentage by surface
  first_serve_by_surface <- atp_clean %>%
    filter(!is.na(winner_1st_serve_pct)) %>%
    group_by(surface) %>%
    summarise(
      avg_first_serve_pct = mean(winner_1st_serve_pct, na.rm = TRUE),
      matches = n()
    ) %>%
    arrange(desc(avg_first_serve_pct))
  
  # Display results
  first_serve_by_surface
  
  # Create visualization
  ggplot(first_serve_by_surface, aes(x = reorder(surface, avg_first_serve_pct), 
                                    y = avg_first_serve_pct, fill = surface)) +
    geom_bar(stat = "identity") +
    labs(title = "Average First Serve Percentage by Surface",
         x = "Surface",
         y = "First Serve %") +
    theme(legend.position = "none")
}
```

# 7. Statistical Analysis {.tabset}

## Surface Impact on Match Time {.tabset}

### Statistical Test Results
```{r surface-duration-test}
# ANOVA test for match duration by surface
if("minutes" %in% colnames(atp_clean) && "surface" %in% colnames(atp_clean)) {
  # Subset data for the test
  duration_data <- atp_clean %>%
    filter(!is.na(minutes))
  
  # Run ANOVA
  duration_anova <- aov(minutes ~ surface, data = duration_data)
  anova_summary <- summary(duration_anova)
  
  # Create boxplot visualization
  ggplot(duration_data, aes(x = surface, y = minutes, fill = surface)) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.1) +
    labs(
      title = "Match Duration by Surface Type",
      subtitle = "Distribution of match lengths across different playing surfaces",
      x = "Surface",
      y = "Duration (minutes)"
    ) +
    theme_minimal() +
    theme(legend.position = "none") +
    scale_fill_viridis_d()

  # Display ANOVA results
  datatable(
    data.frame(
      Metric = c("F-statistic", "p-value", "Significant?"),
      Value = c(
        round(anova_summary[[1]]$`F value`[1], 2),
        format.pval(anova_summary[[1]]$`Pr(>F)`[1], digits = 3),
        ifelse(anova_summary[[1]]$`Pr(>F)`[1] < 0.05, "Yes", "No")
      )
    ),
    options = list(dom = 't', ordering = FALSE),
    caption = "ANOVA Results: Impact of Surface on Match Duration"
  )
  
  # Calculate mean durations for interpretation
  surface_means <- duration_data %>%
    group_by(surface) %>%
    summarise(
      mean_duration = mean(minutes, na.rm = TRUE),
      sd_duration = sd(minutes, na.rm = TRUE)
    ) %>%
    arrange(desc(mean_duration))
  
  longest_surface <- surface_means$surface[1]
  shortest_surface <- surface_means$surface[nrow(surface_means)]
  mean_diff <- round(surface_means$mean_duration[1] - surface_means$mean_duration[nrow(surface_means)], 1)
}
```

### Interpretation
The analysis of match duration across different surfaces reveals several key findings:

1. **Statistical Significance**: The ANOVA test shows whether there are significant differences in match duration across surfaces. A p-value less than 0.05 indicates that surface type has a significant impact on match length.

2. **Practical Impact**: On average, matches on `r longest_surface` tend to last `r mean_diff` minutes longer than matches on `r shortest_surface`. This substantial difference highlights how surface characteristics influence game play and match duration.

3. **Variability**: The boxplot visualization shows the spread of match durations for each surface, with overlapping distributions indicating that while surface type influences match length, other factors also play important roles.

## First Serve Performance Analysis {.tabset}

### Statistical Test Results
```{r serve-analysis}
# Analysis of first serve percentage and match outcome
serve_data <- atp_clean %>%
  filter(!is.na(winner_1st_serve_pct) & !is.na(loser_1st_serve_pct)) %>%
  mutate(
    serve_diff = winner_1st_serve_pct - loser_1st_serve_pct
  )

# Create density plot
ggplot(serve_data, aes(x = serve_diff)) +
  geom_density(fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of First Serve Percentage Difference",
    subtitle = "Difference between Winners and Losers",
    x = "Winner's First Serve % - Loser's First Serve %",
    y = "Density"
  ) +
  theme_minimal()

# Paired t-test
t_test_result <- t.test(serve_data$winner_1st_serve_pct, 
                        serve_data$loser_1st_serve_pct, 
                        paired = TRUE)

# Display results
datatable(
  data.frame(
    Metric = c("Mean Difference", "t-statistic", "p-value", "95% CI Lower", "95% CI Upper"),
    Value = c(
      round(t_test_result$estimate, 2),
      round(t_test_result$statistic, 2),
      format.pval(t_test_result$p.value, digits = 3),
      round(t_test_result$conf.int[1], 2),
      round(t_test_result$conf.int[2], 2)
    )
  ),
  options = list(dom = 't', ordering = FALSE),
  caption = "Paired t-test Results: Winner vs. Loser First Serve %"
)

# Calculate key metrics for interpretation
mean_diff <- round(t_test_result$estimate, 1)
```

### Interpretation
The analysis of first serve performance reveals:

1. **Statistical Significance**: The paired t-test examines whether winners consistently have different first serve percentages compared to losers. The extremely low p-value indicates this difference is not due to chance.

2. **Effect Size**: Winners typically have a first serve percentage that is `r mean_diff`% higher than losers. This difference, while seemingly small, can be crucial in high-level tennis where margins are tight.

3. **Practical Impact**: The density plot shows a clear shift toward positive values, indicating that winners generally maintain better first serve percentages. This suggests that consistent first serve performance is a key factor in match outcomes.

## Handedness Analysis {.tabset}

### Match Outcomes by Handedness
```{r handedness-analysis}
# Create contingency table
handedness_data <- atp_clean %>%
  filter(!is.na(winner_hand) & !is.na(loser_hand)) %>%
  mutate(
    matchup = paste(winner_hand, "vs", loser_hand)
  )

# Create stacked bar chart
ggplot(handedness_data, aes(x = winner_hand, fill = loser_hand)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d() +
  labs(
    title = "Match Outcomes by Player Handedness",
    subtitle = "Distribution of winner-loser handedness combinations",
    x = "Winner's Hand",
    y = "Proportion",
    fill = "Loser's Hand"
  ) +
  theme_minimal()

# Chi-square test
chi_result <- chisq.test(table(handedness_data$winner_hand, handedness_data$loser_hand))

# Calculate proportions for interpretation
hand_props <- handedness_data %>%
  group_by(winner_hand, loser_hand) %>%
  summarise(n = n()) %>%
  group_by(winner_hand) %>%
  mutate(prop = n/sum(n))

# Display results
datatable(
  data.frame(
    Metric = c("Chi-square statistic", "p-value", "Significant?"),
    Value = c(
      round(chi_result$statistic, 2),
      format.pval(chi_result$p.value, digits = 3),
      ifelse(chi_result$p.value < 0.05, "Yes", "No")
    )
  ),
  options = list(dom = 't', ordering = FALSE),
  caption = "Chi-square Test Results: Association between Winner and Loser Handedness"
)
```

### Interpretation
The analysis of player handedness reveals:

1. **Overall Pattern**: The stacked bar chart shows the proportion of matches won by right-handed and left-handed players against opponents of each handedness type. This visualization helps identify any advantages or disadvantages based on handedness matchups.

2. **Statistical Significance**: The chi-square test examines whether there's a significant association between winner and loser handedness. A p-value less than 0.05 would suggest that certain handedness matchups are more likely to result in wins than others.

3. **Practical Implications**: The proportions show whether left-handed players have any advantage over right-handed players, or vice versa. This information could be valuable for players and coaches in preparing for matches against opponents with different handedness.

# 8. Interesting Findings {.tabset}

## Key Insights
```{r interesting-findings}
# Find upset matches (lower ranked player beating higher ranked player)
if(all(c("winner_rank", "loser_rank") %in% colnames(atp_clean))) {
  upsets <- atp_clean %>%
    filter(!is.na(winner_rank) & !is.na(loser_rank)) %>%
    filter(winner_rank > loser_rank) %>%
    mutate(
      rank_difference = winner_rank - loser_rank,
      upset_magnitude = rank_difference / loser_rank
    ) %>%
    arrange(desc(upset_magnitude))
  
  # Show top upsets with DT
  top_upsets <- head(upsets[, c("winner_name", "winner_rank", "loser_name", "loser_rank", 
                               "rank_difference", "tourney_name")], 5)
  
  datatable(top_upsets,
            caption = "Top 5 Biggest Upsets (by ranking difference magnitude)",
            options = list(dom = "t", ordering = FALSE))
}
```

## Longest Matches
```{r longest-matches}
# Find longest matches
if("minutes" %in% colnames(atp_clean)) {
  longest_matches <- atp_clean %>%
    filter(!is.na(minutes)) %>%
    arrange(desc(minutes))
  
  # Show longest matches with DT
  top_longest <- head(longest_matches[, c("winner_name", "loser_name", "minutes", 
                                         "tourney_name", "round")], 5)
  
  datatable(top_longest,
            caption = "Top 5 Longest Matches",
            options = list(dom = 't', ordering = FALSE))
}
```

# 9. Conclusion {.tabset}

## Summary
```{r summary-findings}
# Calculate key summary statistics
summary_stats <- data.frame(
  Metric = c(
    "Total Matches", 
    "Number of Tournaments",
    "Number of Different Winners",
    "Average Match Duration (mins)",
    "Most Common Surface",
    "Average First Serve Percentage",
    "Average Aces per Match",
    "Grand Slam Matches",
    "ATP 500 Matches",
    "ATP 250 Matches",
    "Masters 1000 Matches",
    "Davis Cup Matches",
    "Upset Rate (%)"
  ),
  Value = c(
    nrow(atp_clean),
    length(unique(atp_clean$tourney_id)),
    length(unique(atp_clean$winner_name)),
    round(mean(atp_clean$minutes, na.rm = TRUE), 1),
    names(which.max(table(atp_clean$surface))),
    round(mean(atp_clean$winner_1st_serve_pct, na.rm = TRUE), 1),
    round(mean(atp_clean$w_ace + atp_clean$l_ace, na.rm = TRUE), 1),
    sum(atp_clean$tourney_level_detail == "Grand Slam", na.rm = TRUE),
    sum(atp_clean$tourney_level_detail == "ATP Tour 500", na.rm = TRUE),
    sum(atp_clean$tourney_level_detail == "ATP Tour 250", na.rm = TRUE),
    sum(atp_clean$tourney_level_detail == "Masters 1000", na.rm = TRUE),
    sum(atp_clean$tourney_level_detail == "Davis Cup", na.rm = TRUE),
    round(100 * sum(atp_clean$winner_rank > atp_clean$loser_rank, na.rm = TRUE) / 
            sum(!is.na(atp_clean$winner_rank) & !is.na(atp_clean$loser_rank)), 1)
  )
)

# Display summary using DT
datatable(summary_stats, 
          caption = "Summary of Key Tennis Statistics",
          options = list(dom = 't', 
                         ordering = FALSE,
                         pageLength = nrow(summary_stats)))
```

## Future Work
To complete this analysis for the final presentation, I will:

1. Develop more sophisticated statistical models
2. Create polished, consistent visualizations 
3. Prepare a cohesive narrative around the key findings
4. Explore adding the racquet data, as mentioned in my proposal
